\Large{Adversarial Variational Optimization of Non-Differentiable Simulators}

### 2017-07-19 tex/paper.tex
save

Under mild restrictions outlined in  \citep{2012arXiv1212.4507S}, the bound
$U(\bfpsi)$ is differentiable with respect to $\bfpsi$, and using the log-likelihood
trick its gradient can be rewritten as:
\begin{align}\label{eqn:approx-grad}
    \nabla_\bfpsi U(\bfpsi) &= \nabla_\bfpsi \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [f(\bftheta)] \nonumber \\
    %&= \nabla_\bfpsi \int f(\bftheta)  q(\bftheta|\bfpsi)  d\bftheta \nonumber \\
 %   &= \int f(\bftheta) \nabla_\bfpsi q(\bftheta|\bfpsi)  d\bftheta \nonumber \\
   % &= \int \left[ f(\bftheta) \nabla_\bfpsi \log q(\bftheta|\bfpsi) \right]  q(\bftheta|\bfpsi) d\bftheta \nonumber \\
    &= \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [f(\bftheta) \nabla_\bfpsi \log q(\bftheta|\bfpsi)]
\end{align}
Effectively, this means that provided that the score function $\nabla_\bfpsi \log
q(\bftheta|\bfpsi)$ of the proposal is known and that one can evaluate
$f(\mathbf{\bftheta})$ for any $\bftheta$, then one can construct empirical
estimates of Eqn.~\ref{eqn:approx-grad}, which can in turn be used to minimize
$U(\bfpsi)$ with stochastic gradient descent. We will use
Adam~\cite{2014arXiv1412.6980K}, but note the opportunity to make  use instead of the Natural Evolution Strategy
algorithm~\citep{2011arXiv1106.4487W} or of variance reduction techniques~\citep{ranganath2014black}.
%Algorithm~\ref{alg:avo} represents the simplest version of AVO; however, the variance of the noisy estimator of the gradients may be too large to be useful in many problems. Variance reduction techniques such as Rao-Blackwellization and Control Variates can be applied here as they are in black box variational inference~\citep{ranganath2014black}.

@@

Algorithm~\ref{alg:avo} represents the simplest version of AVO; however, the
variance of the noisy estimator of the gradients may be too large to be useful in many problems. Improvements are possible using variance reduction techniques such as Rao-Blackwellization,
Control Variates, or quasi-Monte Carlo~\citep{ranganath2014black, tran2017variational}.
Similarly, various fast algorithms for calculating the exact (entropically regularized) Wasserstein distance on
empirical distributions are being developed~\citep{cuturi2013sinkhorn, genevay2016stochastic, montavon2016wasserstein}.
%Moreover, the algorithm could be adapted to use estimates to the KL distance~\citep{perez2009estimation} without the adversarial optimization for the critic.
These variants may be particularly useful for small $N$.

@@

When $\rho$ is the Wasserstein distance, $\bftheta^*$ is referred to as the
minimum Wasserstein estimator (MWE). When the model is well specified, $\bftheta^*$
coincides with the true data-generating parameter; however, if the model is
misspecified, $\bftheta^*$ is typically different from the maximum likelihood
estimator (MLE), which is the minimizer of $KL(p_r(\bfx) | p(\bfx |
\bftheta))$~\citep{bernton2017inference}. In order to more effectively target
point estimates $\bftheta^*$, we augment Eqn.~\ref{eqn:vo-ug} with a
regularization term corresponding to the differential entropy $H$ of the
proposal distribution $q(\bftheta | \bfpsi)$. That is,
\begin{equation}
    U_g = \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [ {\cal L}_g ] + \gamma H(q(\bftheta|\bfpsi))
\end{equation}
where $\gamma \in \mathbb{R}^+$ is a hyper-parameter controlling the trade-off
between the generator objective and the tightness of the proposal distribution.
We refer to the algorithm with $\gamma>0$ as AVO-MWE.
For small values of $\gamma$,
proposal distributions with large entropy are not penalized, which results
in learning a smeared variation of the original simulator corresponding to the posterior predictive model. On the other hand,
for large values of $\gamma$, the procedure is encouraged to fit a proposal
distribution with low entropy, which has the effect of concentrating its density
tightly around one or a few $\bftheta$ values.

@@

Perhaps the closest to our work is~\cite{bernton2017inference}, which uses the
Wasserstein distance both to find point estimates $\bftheta^*$ and as a part of
a rejection sampler in an ABC-like setup (as opposed to variational inference).
They emphasize that this approach allows one to eliminate the summary statistics
typically used in ABC and calculate the Wasserstein distance explicitly, without
making use of the Kantorovich-Rubinstein duality and of a critic $d$. For high-dimensional
data, they note that this is computationally expensive and introduce an
approximation based on projection of the data onto Hilbert space-filling curves.
Their Remark 5.1 points out that~\citep{montavon2016wasserstein} proposed an
approximation of the gradient of an entropy-regularized Wasserstein distance,
which uses a similar duality. They note that ``unfortunately, it is not
applicable in the setting of purely generative models, as it involves point-wise
evaluations of the derivative of the log-likelihood.'' Thus, our contribution is
to provide gradients of an approximate MWE by taking expectations with the
variational program $q(\bfx | \bfpsi)$. This paired with the critic
Kantorovich-Rubinstein dual formulation of the Wasserstein distance
allows us to work in high dimensions without summary statistics and to
optimize $\bfpsi$ with stochastic gradient descent. Additionally,
our AVO-EB procedure avoids the inefficiencies of
their Wasserstein rejection sampler incurred from using the ABC-likelihood.

>>>>

Under mild restrictions outlined in  \citep{2012arXiv1212.4507S}, the bound
$U(\bfpsi)$ is differentiable with respect to $\bfpsi$, and using the log-likelihood
trick its gradient can be rewritten as:
\begin{align}\label{eqn:approx-grad}
    \nabla_\bfpsi U(\bfpsi) &= \nabla_\bfpsi \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [f(\bftheta)] \nonumber \\
    %&= \nabla_\bfpsi \int f(\bftheta)  q(\bftheta|\bfpsi)  d\bftheta \nonumber \\
 %   &= \int f(\bftheta) \nabla_\bfpsi q(\bftheta|\bfpsi)  d\bftheta \nonumber \\
   % &= \int \left[ f(\bftheta) \nabla_\bfpsi \log q(\bftheta|\bfpsi) \right]  q(\bftheta|\bfpsi) d\bftheta \nonumber \\
    &= \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [f(\bftheta) \nabla_\bfpsi \log q(\bftheta|\bfpsi)]
\end{align}
Effectively, this means that provided that the score function $\nabla_\bfpsi \log
q(\bftheta|\bfpsi)$ of the proposal is known and that one can evaluate
$f(\mathbf{\bftheta})$ for any $\bftheta$, then one can construct empirical
estimates of Eqn.~\ref{eqn:approx-grad}, which can in turn be used to minimize
$U(\bfpsi)$ with stochastic gradient descent (or a variant thereof, robust to noise
and parameter scaling).
% We will use
% Adam~\cite{2014arXiv1412.6980K}, but note the opportunity to make  use instead of the Natural Evolution Strategy
% algorithm~\citep{2011arXiv1106.4487W} or of variance reduction techniques
% such as Rao-Blackwellization,
% Control Variates, or quasi-Monte Carlo~\citep{ranganath2014black, tran2017variational}.
%Algorithm~\ref{alg:avo} represents the simplest version of AVO; however, the variance of the noisy estimator of the
%gradients may be too large to be useful in many problems. Variance reduction techniques such as Rao-Blackwellization and Control Variates can be applied here as they are in black box variational inference~\citep{ranganath2014black}.

@@

% Algorithm~\ref{alg:avo} represents the simplest version of AVO; however, the
% variance of the noisy estimator of the gradients may be too large to be useful in many problems.
% Improvements are possible using variance reduction techniques such as Rao-Blackwellization,
% Control Variates, or quasi-Monte Carlo~\citep{ranganath2014black, tran2017variational}.
% Similarly, various fast algorithms for calculating the exact (entropically regularized) Wasserstein distance on
% empirical distributions are being developed~\citep{cuturi2013sinkhorn, genevay2016stochastic, montavon2016wasserstein}.
% %Moreover, the algorithm could be adapted to use estimates to the KL distance~\citep{perez2009estimation} without the adversarial optimization for the critic.
% These variants may be particularly useful for small $N$.

Algorithm~\ref{alg:avo} represents the simplest version of AVO; however, the
variance of the noisy estimator of the gradients may be too large to be useful in many problems.
We use Adam~\cite{2014arXiv1412.6980K}, but note the opportunity use instead the Natural Evolution Strategy
algorithm~\citep{2011arXiv1106.4487W} or variance reduction techniques
such as Rao-Blackwellization,
Control Variates, or quasi-Monte Carlo~\citep{ranganath2014black, tran2017variational}.

@@

When $\rho$ is the Wasserstein distance, $\bftheta^*$ is referred to as the
minimum Wasserstein estimator (MWE). When the model is well specified, $\bftheta^*$
coincides with the true data-generating parameter; however, if the model is
misspecified, $\bftheta^*$ is typically different from the maximum likelihood
estimator (MLE), which is the minimizer of $\text{KL}(p_r(\bfx) \parallel p(\bfx |
\bftheta))$~\citep{bernton2017inference}. In order to more effectively target
point estimates $\bftheta^*$, we augment Eqn.~\ref{eqn:vo-ug} with a
regularization term corresponding to the differential entropy $H$ of the
proposal distribution $q(\bftheta | \bfpsi)$. That is,
\begin{equation}
    U_g = \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [ {\cal L}_g ] + \gamma H(q(\bftheta|\bfpsi))
\end{equation}
where $\gamma \in \mathbb{R}^+$ is a hyper-parameter controlling the trade-off
between the generator objective and the tightness of the proposal distribution.
We refer to the algorithm with $\gamma>0$ as AVO-MWE.

For small values of $\gamma$,
proposal distributions with large entropy are not penalized, which results
in learning a smeared variation of the original simulator corresponding to the posterior predictive model. On the other hand,
for large values of $\gamma$, the procedure is encouraged to fit a proposal
distribution with low entropy, which has the effect of concentrating its density
tightly around one or a few $\bftheta$ values.

@@

Perhaps the closest to our work is~\cite{bernton2017inference}, which uses the
Wasserstein distance both to find point estimates $\bftheta^*$ and as a part of
a rejection sampler in an ABC-like setup (as opposed to variational inference).
They emphasize that this approach allows one to eliminate the summary statistics
typically used in ABC and calculate the Wasserstein distance explicitly, without
making use of the Kantorovich-Rubinstein duality and of a critic $d$. For high-dimensional
data, they note that this is computationally expensive and introduce an
approximation based on projection of the data onto Hilbert space-filling curves.
Their Remark 5.1 points out that~\citep{montavon2016wasserstein} proposed an
approximation of the gradient of an entropy-regularized Wasserstein distance,
which uses a similar duality. They note that ``unfortunately, it is not
applicable in the setting of purely generative models, as it involves point-wise
evaluations of the derivative of the log-likelihood.'' Thus, our contribution is
to provide gradients of an approximate MWE by taking expectations with the
variational program $q(\bfx | \bfpsi)$. This paired with the critic
Kantorovich-Rubinstein dual formulation of the Wasserstein distance
allows us to work in high dimensions without summary statistics and to
optimize $\bfpsi$ with stochastic gradient descent. Additionally,
our AVO-EB procedure avoids the inefficiencies of
their Wasserstein rejection sampler incurred from using the ABC-likelihood.
In the case of small $N$, we note however that fast algorithms for calculating the exact (entropically regularized) Wasserstein distance on
empirical distributions~\citep{cuturi2013sinkhorn, genevay2016stochastic, montavon2016wasserstein}
are alternatives to adversarial learning that are worth considering.

% Similarly, various fast algorithms for calculating the exact (entropically regularized) Wasserstein distance on
% empirical distributions are being developed~\citep{cuturi2013sinkhorn, genevay2016stochastic, montavon2016wasserstein}.
% %Moreover, the algorithm could be adapted to use estimates to the KL distance~\citep{perez2009estimation} without the adversarial optimization for the critic.
% These variants may be particularly useful for small $N$.
